---
title: "Background Ion Fragmentation Audit (Thermo .raw + rawrr)"
author: "Metabolomics Utilities"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
params:
  raw_files: !r character()
  mz_tolerance: 0.005
  top_n: 20
  min_precursor_mz: 80
  max_precursor_mz: 1500
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 6
)
```

## Overview

This report scans Thermo Scientific `.raw` files with **rawrr** and identifies precursor masses that were fragmented repeatedly.
These recurring precursor m/z values are commonly background ions, contaminants, or persistent chemical noise.

The report produces:

1. A ranked table of the top recurring fragmented precursor m/z values.
2. A bar plot showing how often each top precursor was fragmented.
3. Fragmentation spectra (MS2) for those top precursor m/z groups.

## Requirements

```{r requirements}
required_pkgs <- c("rawrr", "dplyr", "purrr", "tidyr", "ggplot2", "stringr", "forcats", "scales")
missing_pkgs <- required_pkgs[!vapply(required_pkgs, requireNamespace, quietly = TRUE, FUN.VALUE = logical(1))]

if (length(missing_pkgs) > 0) {
  stop(
    "Missing required packages: ", paste(missing_pkgs, collapse = ", "),
    "\nInstall them before rendering this report."
  )
}
```

## Input files

Provide input files either:

- via `params$raw_files` when rendering, or
- by editing the `manual_raw_files` vector below.

```{r input-files}
manual_raw_files <- character()

raw_files <- unique(c(params$raw_files, manual_raw_files))
raw_files <- raw_files[file.exists(raw_files)]

if (length(raw_files) == 0) {
  stop("No existing .raw files found. Provide paths in params$raw_files or manual_raw_files.")
}

raw_files
```

## Helper functions

```{r helpers}
library(rawrr)
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)
library(stringr)
library(forcats)
library(scales)

extract_ms2_headers <- function(raw_file, min_mz, max_mz) {
  idx <- rawrr::readIndex(raw_file)
  ms2_scans <- idx$scan[idx$msOrder == "Ms2"]

  if (length(ms2_scans) == 0) {
    return(tibble())
  }

  hdr <- rawrr::readTrailer(raw_file, scan = ms2_scans)

  precursor_col <- names(hdr)[str_detect(names(hdr), regex("Monoisotopic M/Z|Selected Ion m/z|MS2 Isolation", ignore_case = TRUE))]
  if (length(precursor_col) == 0) {
    stop("Could not identify precursor m/z column in trailer metadata for file: ", raw_file)
  }
  precursor_col <- precursor_col[[1]]

  tibble(
    file = raw_file,
    scan = ms2_scans,
    precursor_mz = suppressWarnings(as.numeric(hdr[[precursor_col]]))
  ) %>%
    filter(!is.na(precursor_mz)) %>%
    filter(precursor_mz >= min_mz, precursor_mz <= max_mz)
}

bin_precursor_mz <- function(x, tolerance) {
  round(x / tolerance) * tolerance
}

extract_spectrum_peaks <- function(raw_file, scan_no, precursor_bin, top_k = 300) {
  sp <- rawrr::readSpectrum(raw_file, scan = scan_no)

  tibble(
    precursor_bin = precursor_bin,
    file = raw_file,
    scan = scan_no,
    mz = sp[[1]]$mz,
    intensity = sp[[1]]$intensities
  ) %>%
    filter(!is.na(mz), !is.na(intensity), intensity > 0) %>%
    arrange(desc(intensity)) %>%
    slice_head(n = top_k)
}

estimate_scan_tic <- function(raw_file, scan_no) {
  sp <- rawrr::readSpectrum(raw_file, scan = scan_no)
  sum(sp[[1]]$intensities, na.rm = TRUE)
}
```

## Collect MS2 precursor events

```{r collect-events}
precursor_events <- map_dfr(
  raw_files,
  extract_ms2_headers,
  min_mz = params$min_precursor_mz,
  max_mz = params$max_precursor_mz
)

if (nrow(precursor_events) == 0) {
  stop("No MS2 precursor events found in the provided files.")
}

precursor_events <- precursor_events %>%
  mutate(precursor_bin = bin_precursor_mz(precursor_mz, params$mz_tolerance))

precursor_summary <- precursor_events %>%
  count(precursor_bin, name = "n_fragmented") %>%
  arrange(desc(n_fragmented)) %>%
  slice_head(n = params$top_n)

precursor_summary
```

## Top recurring fragmented precursor masses

```{r barplot-top-precursors}
plot_df <- precursor_summary %>%
  mutate(precursor_label = sprintf("%.4f", precursor_bin)) %>%
  mutate(precursor_label = fct_reorder(precursor_label, n_fragmented))

ggplot(plot_df, aes(x = precursor_label, y = n_fragmented)) +
  geom_col(fill = "#2C7FB8") +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(
    title = sprintf("Top %d repeatedly fragmented precursor masses", params$top_n),
    x = "Precursor m/z (binned)",
    y = "Number of MS2 fragmentations"
  ) +
  theme_minimal(base_size = 12)
```

## Fragmentation spectra of top recurring precursor masses

For each precursor m/z bin, one representative MS2 scan with the highest total ion current is selected and visualized.

```{r spectra-top-precursors}
top_precursor_events <- precursor_events %>%
  semi_join(precursor_summary, by = "precursor_bin")

representative_scans <- top_precursor_events %>%
  mutate(scan_tic = map2_dbl(file, scan, estimate_scan_tic)) %>%
  group_by(precursor_bin) %>%
  slice_max(order_by = scan_tic, n = 1, with_ties = FALSE) %>%
  ungroup()

spectra_long <- pmap_dfr(
  list(representative_scans$file, representative_scans$scan, representative_scans$precursor_bin),
  extract_spectrum_peaks
)

if (nrow(spectra_long) == 0) {
  stop("No spectral peaks extracted for representative MS2 scans.")
}

spectra_plot_df <- spectra_long %>%
  mutate(precursor_label = sprintf("%.4f", precursor_bin)) %>%
  group_by(precursor_label) %>%
  mutate(rel_intensity = intensity / max(intensity, na.rm = TRUE) * 100) %>%
  ungroup()

ggplot(spectra_plot_df, aes(x = mz, y = rel_intensity)) +
  geom_segment(aes(xend = mz, yend = 0), linewidth = 0.25, color = "#D95F0E") +
  facet_wrap(~ precursor_label, scales = "free_x", ncol = 4) +
  labs(
    title = "Representative MS2 fragmentation spectra for top recurring precursor masses",
    x = "Fragment m/z",
    y = "Relative intensity (%)"
  ) +
  theme_minimal(base_size = 11)
```

## Optional export

```{r export-summary}
# write.csv(precursor_summary, "top_background_precursors.csv", row.names = FALSE)
```

## Rendering example

Render from shell:

```bash
Rscript -e 'rmarkdown::render(
  "rawrr_background_ions_report.Rmd",
  params = list(
    raw_files = c("/path/to/file1.raw", "/path/to/file2.raw"),
    mz_tolerance = 0.005,
    top_n = 20
  )
)'
```
